Utilizing Parallel Workers: LLVM's Vectorization Plan

### 2025-01-26 paper/paper.tex
paper

Uncountable loops can typically also not be vectorized due to their unknown number of iterations.

Finally, some instructions, such as integer division, do inherently not have a vectorized instruction equivalent.

@@

Some vector instructions can be costly on most CPUs. Examples are gather, scatter, shuffle,
and permutation instructions. control flow with many non-adjacent memory loads/stores or 
operations that combine different vector elements, such as reductions, can decrease vectorization 
performance.

Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow often includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the Vectorization Factor, a scalar
epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
to overlapping memory regions, in which case the scalar version of the loop is executed.

@@

\paragraph{H-CFG}
This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
control-flow graph and its components are the main elements of a single VPlan.
The nodes of this H-CFG are one of three types:

@@

Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
shows the loop part of the H-CFG of a VPlan that models the vectorization approach described
in Figure~\ref{fig:inner-loop-vec}.

@@

There are many different types of \texttt{VPRecipes}, most notably:
\begin{itemize}
  \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
  or multiple times (\texttt{REPLICATE}) into the vectorized IR when executed.
  \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
  vector-equivalent when executed.
  \item \texttt{VPInstruction} has no underlying IR instruction it is based on and is used 
  to model newly needed instructions for the vectorized IR. When executed it will 
  \texttt{EMIT} one or more new IR instructions.
\end{itemize}

@@

\subsection{State of VPlan (LLVM 19.1.4)}
\emergencystretch 6em
In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
used to model, optimize, and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
While outer loop vectorization via VPlan has already seen some development, any 
outer loop vectorization features are still experimental and must be explicitly enabled 
by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
feature flag~\cite{llvmouterloop,llvmouterloopstatus}.

@@

\subsubsection{Phase 1: Legality}
The main part of this phase is checking whether it is even legal to optimize a given loop. 
In addition to the general vectorization legality concepts mentioned in 
Section~\ref{sec:vec-legal}, LV has a several extra conditions to check.

@@

\subsubsection{Phase 2: Planning}
After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
A VPlan is initialized by creating the H-CFG skeleton building blocks like preheaders, a 
region block for the vector loop, a middle block with the potential runtime check for the scalar 
epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
is iterated in topological order, and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
After modeling each scalar IR instruction with a \texttt{VPRecipe}, several VPlan transformations 
are executed, optimizing the initialized VPlan.

@@

\paragraph{Cost Model}
Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instruction are then generally just lookup tables influenced by the underlying 
compile target.

\subsubsection{Phase 3: Execution}
After the best VPlan has been chosen, it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
For this VPlan, some final transformations are run depending on a specifically chosen VF and UF.
First, a vectorized code skeleton is created in the IR, and then the VPlan is executed. 
Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which executes all 
of its \texttt{VPRecipes} in order.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.35\textwidth]{images/inner-loop-vec-loop-cfg-color.png}
  \caption{The generalized CFG for a vectorized loop with trip count \texttt{N}, 
  iteration variable \texttt{i}, vectorization factor \texttt{VF}, 
  and unroll factor \texttt{UF} in LLVM. A VPlan would explicitly model the red-highlighted 
  region in its H-CFG.}
  \label{fig:inner-loop-vec-cfg}
\end{figure}

@@

\subsection{Future of VPlan}
Many of the VPlan internals and surrounding systems are still in active development. To stay up-to-date 
with the latest developments regarding auto-vectorization in LLVM, one can follow the 
Community Forum\footnote{\url{https://discourse.llvm.org/c/ir-optimizations/loop-optimizations/62}} or
attend the monthly online 
meetings\footnote{\url{https://discourse.llvm.org/t/monthly-vectorizer-online-sync-up/78978/3}}.

\subsubsection{VPlan and Integration with LV}
Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
few years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
abstracted scalar IR instructions.

\subsubsection{Outer Loop Vectorization}
Development around outer loop vectorization has slowed as the focus shifted to VPlan internals 
and integration into LV. The inner and outer loop paths are still separated, and the 
experimental \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} flag must be 
set to activate any outer 
loop vectorization. Further development of the outer loop path and eventually merging it 
with the innermost loop path is planned.

\subsubsection{Merging with SLP Vectorize}
Merging the existing SLP Vectorizer with the current VPlan infrastructure is in the early planning phase.

\subsubsection{Function Vectorization}
There is currently no active development for whole-function vectorization in the
llvm-project's upstream.

@@

@@

>>>>

Uncountable loops, meaning loops where the total iteration count cannot be determined before the 
loop is executed, can often not be vectorized. This is because their exit condition usually depends on 
the loop's body itself, making it impossible to predetermine how many vector lanes are needed for the next 
loop iteration when executing.

Finally, some instructions do generally not have a direct vectorized 
instruction equivalent. For example Integer division, commonly implemented using the 
\textit{Newton-Raphson method}, is already a complex and iterative operation 
on modern CPUs. Applying this algorithm 
to multiple data elements in a single vector register would result in even more implementation complexity 
and is thus considered not worthwhile in most cases.
Workarounds for some specialized cases do exist~\cite{vecintdiv}.

@@

The actual performance of a vector instruction also heavily depends on the ISA and its 
hardware implementation. 
For example, gather, scatter, 
shuffle, and permutation instructions are typically slow. Control flow with many 
non-adjacent memory accesses or 
operations that combine different vector elements from the same register (horizontal reductions) can thus 
decrease vectorization performance.

Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow often includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the Vectorization Factor, a scalar
epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
to overlapping memory regions, in which case the scalar version of the loop must be executed due to 
unclear data dependencies.

@@

\paragraph{H-CFG}
This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
control-flow graph and its components are the main elements of a single VPlan. It is hierarchical
in that a single node of this graph can be a complete subgraph, representing more complex nested
behavior.
Specifically, any node of this H-CFG is one of three types:

@@

Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
shows the loop part of the H-CFG of a VPlan that models the vectorization approach described
in Figure~\ref{fig:inner-loop-vec}.
% TODO: Explain contents of both figures better

@@

There are many different types of \texttt{VPRecipes}, most notably:
\begin{itemize}
  \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
  or VF * UF times (\texttt{REPLICATE}) into the vectorized IR when executed.
  \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
  vector-equivalent when executed.
  \item \texttt{VPInstruction} has no underlying IR instruction it is based on and is used 
  to model newly needed instructions for the vectorized IR. When executed it will 
  \texttt{EMIT} one or more new IR instructions.
  % TODO: Explain some instructions more in detail (e.g. CANONICAL-INDUCTION, SCALAR-STEPS)
\end{itemize}

@@

% TODO: VPlan Transformations example

\section{State of VPlan (LLVM 19.1.4)}
\emergencystretch 6em
In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
used to model, optimize, and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
While outer loop vectorization via VPlan has already seen some development, any 
outer loop vectorization features are still experimental and must be explicitly enabled 
by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
feature flag~\cite{llvmouterloop,llvmouterloopstatus}.

@@

\subsection{Phase 1: Legality}
The main part of this phase is checking whether it is even legal to optimize a given loop. 
In addition to the general vectorization legality concepts mentioned in 
Section~\ref{sec:vec-legal}, LV has a several extra conditions to check.

@@

\subsection{Phase 2: Planning}
% TODO: Connect better to previous phase (expand a bit on how the legality artifacts are used when planning)
After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
A VPlan is initialized by creating the H-CFG skeleton building blocks like preheaders, a 
region block for the vector loop, a middle block with the potential runtime check for the scalar 
epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
is iterated in topological order, and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
After modeling each scalar IR instruction with a \texttt{VPRecipe}, several VPlan transformations 
are executed, optimizing the initialized VPlan.

@@

\paragraph{Cost Model}
Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instruction are then generally just lookup tables influenced by the underlying 
compile target.
% TODO: Expand on this and explain a bit more

\subsection{Phase 3: Execution}
After the best VPlan has been chosen, it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
For this VPlan, some final transformations are run depending on a specifically chosen VF and UF.
First, a vectorized code skeleton is created in the IR, and then the VPlan is executed. 
Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which in turn executes all 
of its \texttt{VPRecipes} in order.

% TODO: How exactly is the translation from VP recipe back to LLVM IR working? Where does the target
%       backend come in to decide what actual instruction?

@@

\begin{figure}
  \centering
  \includegraphics[width=0.35\textwidth]{images/inner-loop-vec-loop-cfg-color.png}
  \caption{The generalized CFG for a vectorized loop with trip count \texttt{N}, 
  iteration variable \texttt{i}, vectorization factor \texttt{VF}, 
  and unroll factor \texttt{UF} in LLVM. A VPlan would so far explicitly model the red-highlighted 
  region in its H-CFG. If a node branches, the condition is stated in the body, and the two possible 
  path edges are marked with T (True) and F (False).}
  \label{fig:inner-loop-vec-cfg}
\end{figure}

@@

\section{Future of VPlan}
Many of the VPlan internals and surrounding systems are still in active development. To stay up-to-date 
with the latest developments regarding auto-vectorization in LLVM, one can follow the 
Community Forum\footnote{\url{https://discourse.llvm.org/c/ir-optimizations/loop-optimizations/62}} or
attend the monthly online 
meetings\footnote{\url{https://discourse.llvm.org/t/monthly-vectorizer-online-sync-up/78978/3}}.

\subsection{VPlan and Integration with LV}
Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
few years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
abstracted scalar IR instructions.

\subsection{Outer Loop Vectorization}
Development around outer loop vectorization has slowed as the focus shifted to VPlan internals 
and integration into LV. The inner and outer loop paths are still separated, and the 
experimental \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} flag must be 
set to activate any outer 
loop vectorization. Further development of the outer loop path and eventually merging it 
with the innermost loop path is planned.

\subsection{Merging with SLP Vectorize}
Merging the existing SLP Vectorizer with the current VPlan infrastructure is in the early planning phase.

\subsection{Function Vectorization}
There is currently no active development for whole-function vectorization in the
llvm-project's upstream.

@@

% TODO: Dependency Analysis (SCEV Expressions)
%         - https://www.npopov.com/2023/10/03/LLVM-Scalar-evolution.html
%         - https://www.youtube.com/watch?v=AmjliNp0_00
% TODO: Polly and its use for loop transformations affecting data dependencies
% TODO: Comparisons LLVM with GCC
%         - Tail loop folding (https://github.com/llvm/llvm-project/issues/123069)
%         - Another benchmark: https://www.spec.org/cpu2017/Docs/benchmarks/525.x264_r.html

@@

% TODO: Write about potential future work benchmarking
%       Loop Code Benchmark collection
%       Quick note about collaborative development (example stability & miscompile policy)
%       LLVM Loop Optimization as one of the most active subareas of LLVM development
